---
title: "Final Report"
author: "Joseph Burks"
date: "2024-05-05"
output: github_document
---

# Data Set
The data set is from 538 a news website owned by ABC News with a focus on opinion poll analysis,
economics and politics. The data set utilized is a collection of various demographic and economic information per state, such as percentage of no white citizens, percentage of non citizens and median incomes, as wells as, hate
crime statistics. 

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(factoextra)
hate_crimes <- readr::read_csv("C:/Users/josep/Downloads/hate_crimes.csv")
head(hate_crimes)
```

# Exlporatory Analysis
## Summary Statitics

```{r}
summary_stats <- summarytools::descr(hate_crimes, round.digits = 2, transpose = TRUE)
summarytools::view(summary_stats, method = "render")
```

## Distribution of Variables

```{r}
#ggplot(hate_crimes) +
```

## Hate Crimes by State

```{r,warning =FALSE}
ggplot(hate_crimes) +
  aes(x = reorder(state, avg_hatecrimes_per_100k_fbi), y = avg_hatecrimes_per_100k_fbi) +
  geom_bar(position="dodge",stat="identity",fill="blue",cex=0.75) + 
  coord_flip() +
  labs(title = "Average Hate Crime per 100K by State",
       x = "State", y = "Average Hate Crime per 100K")
```
District of Columbia has by far the largest amount of hate crimes per 100k, more than 
double the second largest of Massachusetts. The states with smallest amount of hate crimes is Wyoming,
Georgia, and Pennsylvania.


## Missing Values 
```{r}
missing.values <- hate_crimes |>
  gather(key = "key", value = "val") |>
  mutate(is.missing = is.na(val)) |>
  group_by(key, is.missing) |>
  summarise(num.missing = n()) |>
  filter(is.missing == T) |>
  select(-is.missing) |>
  arrange(desc(num.missing))


missing.values |>
  ggplot() +
  geom_bar(aes(x=key, y = num.missing), stat = "identity",fill = "orange") +
  labs(x = "variable", y = "number of missing values", title="Number of missing values") +
  theme(axis.text.x = element_text(angle = 45, hjust =1))
```
Luckily the data set does not contain many missing values, and in fact many seems to be the 
result of human error. 538's given source for share of non citizens, Kaiser Family Foundation
does contain the percentage for the non citizens for the missing states, which can be added to the data set.
Similarly all the missing data for hate crimes per 100k from the Southern Poverty Law Center is reported, 
just that in the 4 states with missing data the Southern Poverty Law Center actually reported 0 instincances
of hate crime. Although this is likely due to how there data collection relied on people reporting directly to them.
The last missing data point is a result of Hawaii not sharing hate crime information with the FBI, meaning that the data is not missing completely at random. Since Hawaii is the only state with missing data, we decided to drop the state.
```{r}
no_na_crimes <- hate_crimes
no_na_crimes[20,6] <- 0.01
no_na_crimes[25,6] <- 0.01
no_na_crimes[42,6] <- 0.03

no_na_crimes[is.na(no_na_crimes)] <- 0
no_na_crimes[12,12] <- NA

missing.values2 <- no_na_crimes |>
  gather(key = "key", value = "val") |>
  mutate(is.missing = is.na(val)) |>
  group_by(key, is.missing) |>
  summarise(num.missing = n()) |>
  filter(is.missing == T) |>
  select(-is.missing) |>
  arrange(desc(num.missing))
missing.values2
```
# Principle Component Analysis
```{r}
no_na_crimes <- drop_na(no_na_crimes)
scaled_crimes <- apply(no_na_crimes[,2:12], 2, scale)
crimes_cov <- cov(scaled_crimes)
crimes_eigen <- eigen(crimes_cov)
str(crimes_eigen)
```
According to Kaiser's rule the first three principle components are enough since they are the only ones with eigenvalues greater than 1
```{r}
phi <-crimes_eigen$vectors[,1:3]
phi <- -phi
colnames(phi) <- c("PC1","PC2","PC3")

pc1 <- scaled_crimes %*% phi[,1]
pc2 <- scaled_crimes %*% phi[,2]
pc3 <- scaled_crimes %*% phi[,3]

PC <- data.frame(State = no_na_crimes$state,pc1,pc2,pc3)
head(PC)
```
```{r}
results <- princomp(scaled_crimes)
fviz_pca_biplot(results)
```

\ The biggest contributions for the first principle component are positive contributions from the variables
share_population_with_highschool_degree and median_income, and negative contributions from share_unemployed_seasonal
and share_white_poverty. This indicates that the first principle component is a kind of measure of the states' economy. The biggest contributions for the second component share_non_citizen, share_population_in_metro_areas, and the two hate crime variables. Together the first two principle components explain the nearly 70 percent of the variability in the data.
```{r}
ggplot(PC, aes(pc1, pc2)) + 
  modelr::geom_ref_line(h = 0) +
  modelr::geom_ref_line(v = 0) +
  geom_text(aes(label = State), size = 3) +
  xlab("First Principal Component") + 
  ylab("Second Principal Component") + 
  ggtitle("First Two Principal Components of USArrests Data")
```
Mississippi appears to have above average economy, at least according to the metrics in the data set, and New Hampshire has the worst. California has above average economy and above average hate_crimes, percentage of non citizens and urbanization.

```{r}
PVE <- crimes_eigen$values/sum(crimes_eigen$values)
round(PVE,3)
crimes_eigen$values
```
The first 3 Principle components explain roughly 80 percent of the variability of the data and are the only pcs with eigenvalues greater than 1, so 3 principle components are enough.
